---
title: "Week of 05/23/2022"
author: "Zachary Rowson"
date: "5/26/2022"
output:
  html_document:
        code_folding: hide
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(cache=TRUE)
library(nlme)
library(lattice)
library(data.table)
library(ggplot2)
library(gabi)
```

## Introduction

Here I will attempt to adjust the model for the heterogeneity of variance dependent on the expected value of movement data.

## Fit Models and Evaluate Heterogeneity

```{r, iso-split-data, message=FALSE, warning=FALSE}
load("../data/Padilla_DNT60_pmr0_long.rda")

# Isolate Tebuconazole data
chm <- "Tebuconazole"
DNT60_data <- gabi::data_egids(DNT60pmr0_long)
group <- DNT60_data[cpid==chm, unique(egid)]
Padilla_DNT60_Tebuconazole <- DNT60_data[cpid==chm | (wllt=="v"&egid==group)]

# Split data by experimental phase
data <- copy(Padilla_DNT60_Tebuconazole)
data.L <- Padilla_DNT60_Tebuconazole[t%in%11:30]
data.D <- Padilla_DNT60_Tebuconazole[t%in%31:50]

# Shift time measurements and save concentration as a factor
data.L$t <- data.L$t - mean(data.L$t)
data.L$conc <- as.factor(data.L$conc)
data.D$t <- data.D$t - mean(data.D$t)
data.D$conc <- as.factor(data.D$conc)

# Add well location, inside vs. outside, as co-variate
data.L[rowi%in%c(1,8), well_loc:="outside"][coli%in%c(1,12), well_loc:="outside"]
data.L[is.na(well_loc), well_loc:="inside"]
data.D[rowi%in%c(1,8), well_loc:="outside"][coli%in%c(1,12), well_loc:="outside"]
data.D[is.na(well_loc), well_loc:="inside"]

# Transform data
data.L.sqrt <- copy(data.L)
data.L.sqrt[, sqrty := sqrt(y)]
data.D.sqrt <- copy(data.D)
data.D.sqrt[, sqrty := sqrt(y)]

# Identify individuals to remove
remove <- Padilla_DNT60_Tebuconazole[is.na(y), unique(fishID)]
```
```{r, hier-struct, echo=TRUE}
Light.sqrt <- groupedData(sqrty ~ t | fishID, data = as.data.frame(data.L.sqrt[!(fishID%in%remove)]), order.groups = FALSE)
Dark.sqrt <- groupedData(sqrty ~ t | fishID, data = as.data.frame(data.D.sqrt[!(fishID%in%remove)]), order.groups = FALSE)
```
```{r, aic-func}
normAIC <- function(zz) {
    zz$AIC <- zz$AIC - min(zz$AIC)
    indx <- order(zz$AIC)
    zz[indx,]
}
```

Fit Light phase model.

```{r, fit-Light-model, echo=TRUE, warning=FALSE}
ctrl <- lmeControl(opt = "optim", niterEM = 200, msTol = 1e-10, msMaxIter = 1000, apVar = FALSE)

lmm.L <- lme(sqrty ~ conc*poly(t,degree=2,raw=TRUE),
             data = Light.sqrt,
             random = reStruct(~poly(t,degree=2,raw=TRUE)|fishID,REML=FALSE,pdClass="pdSymm"),
             correlation = corAR1(,form=~t|fishID),
             method = "ML",
             control = ctrl)
```

Fit Dark phase model.

```{r, fit-Dark-model, echo = TRUE, warning=FALSE}
lmm.D <- lme(sqrty ~ conc*poly(t,degree=2,raw=TRUE),
            data = Dark.sqrt,
            random = reStruct(~poly(t,degree=1,raw=TRUE)|fishID,REML=FALSE,pdClass="pdSymm"),
            correlation = corAR1(,form=~t|fishID),
            method = "ML",
            control = ctrl)
```

Visualize residuals by fitted values.

```{r, vis-residuals}
plot(lmm.L, residuals(.) ~ fitted(.), main="Light Phase: Residuals v. Fitted Values Level = 1", abline = 0)
plot(lmm.D, residuals(.) ~ fitted(.), main="Dark Phase: Residuals v. Fitted Values Level = 1", abline = 0)
```

Residuals appear to display a gain-loss like pattern where variance is largest when the expected value is somewhere in the middle of the range of all possible values taken. Add a variance function to evaluate its effects of residuals.

```{r, apply-varFunc, warning=FALSE}
# Transform data
lmm.L.1 <- update(lmm.L, weights=varConstPower())
plot(lmm.L.1, residuals(.) ~ fitted(.), main="Light Phase: Variance Function Added Level = 1", abline = 0)

zz <- AIC(lmm.L, lmm.L.1)
normAIC(zz)
```

Evaluate the normality of within group resiudals.

```{r, qqnorm-L}
qqnorm(lmm.L, ~resid(.), abline=c(0,1), main="Light Model")
qqnorm(lmm.L.1, ~resid(.), abline=c(0,1), main="Light Model with VarFunc")
```

Evaluate qq plots for the Dark phase model.

```{r, qqnorm-D}
qqnorm(lmm.D, ~resid(.), main="Dark Model")
```

Attempt to fit variance function to Dark model.

```{r, varFunc-D}
ctrl <- lmeControl(opt = "optim", niterEM = 200, msTol = 1e-10, msMaxIter = 10000, apVar = FALSE)

lmm.D.1 <- lme(sqrty ~ conc*poly(t,degree=2,raw=TRUE),
             data = Dark.sqrt,
             random = reStruct(~poly(t,degree=2,raw=TRUE)|fishID,REML=FALSE,pdClass="pdSymm"),
             correlation = corAR1(,form=~t|fishID),
             weights = varConstPower(),
             method = "ML",
             control = ctrl)

zz <- AIC(lmm.D, lmm.D.1)
normAIC(zz)
```

```{r, qqnorm-D-var}
qqnorm(lmm.D.1, ~resid(.), main="Dark Model w/ varFunc")
plot(lmm.D.1)
```

Evaluate effect of variance function addition on parameter estimates.

```{r, create-intervals-lmmL}
int.lmm.L <- intervals(lmm.L, which="fixed")
int.lmm.L.1 <- intervals(lmm.L.1, which="fixed")

int.list.L <- lapply(list(int.lmm.L$fixed,int.lmm.L.1$fixed), function(int.df) {
  dt <- as.data.table(int.df)
  concs <- unique(data$conc)
  dt[, conc := c(concs,0,concs,concs[-1])]
  dt[, param := c(rep("B_0",6),"B_1","B_2",rep("B_1",5),rep("B_2",5))]
})

int.dt.L <- do.call('rbind', int.list.L)
int.dt.L[, model := rep(c("No Var Func","Var Func"),each=18)]
```
```{r, plot-intervals-lmmL-lmmLstar}
ggplot(int.dt.L, aes(x=est.,y=log10(conc),color=model)) +
  geom_point() +
  geom_errorbar(aes(xmin=lower, xmax=upper)) +
  facet_wrap(vars(param), scales="free_x") +
  geom_vline(xintercept=0)
```

For Light phase data the effect on parameter estimates is visually apparent.

Evaluate for Dark models.

```{r, create-intervals-lmmD}
int.lmm.D <- intervals(lmm.D, which="fixed")
int.lmm.D.1 <- intervals(lmm.D.1, which="fixed")

int.list.D <- lapply(list(int.lmm.D$fixed,int.lmm.D.1$fixed), function(int.df) {
  dt <- as.data.table(int.df)
  concs <- unique(data$conc)
  dt[, conc := c(concs,0,concs,concs[-1])]
  dt[, param := c(rep("B_0",6),"B_1","B_2",rep("B_1",5),rep("B_2",5))]
})

int.dt.D <- do.call('rbind', int.list.D)
int.dt.D[, model := rep(c("No Var Func","Var Func"),each=18)]
```
```{r, plot-intervals-lmmD-lmmDstar}
ggplot(int.dt.D, aes(x=est.,y=log10(conc),color=model)) +
  geom_point() +
  geom_errorbar(aes(xmin=lower, xmax=upper)) +
  facet_wrap(vars(param), scales="free_x") +
  geom_vline(xintercept=0)
```
 
 For Dark data estimates don't really change.
