---
title: "Using Linear Mixed Effects Models to Model Zebrafish LMR Data"
author: "Zachary Rowson"
date: "3/15/2022"
output:
  html_document: default
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(nlme)
library(data.table)
library(ggplot2)
```

## Introduction

LMEs allow us to model the deviation of individual Zebrafish tracks from the sample averaged parameters. How do we want to model these deviations?

```{r, echo = FALSE}
load("Padilla_DNT60_Loperamide.rda")

rmv <- Padilla_DNT60_Loperamide[is.na(y), unique(fishID)]
dataL <- Padilla_DNT60_Loperamide[t%in%11:30 & !(fishID%in%rmv)]
nestLop.L <- groupedData(y ~ conc|fishID, data =  dataL[conc==0], order.groups = FALSE)

# Set beginning of light to t=0. Set concentration as a factor
nestLop.L$t <- nestLop.L$t - min(nestLop.L$t)
nestLop.L$conc <- as.factor(nestLop.L$conc)
```


## Visualize Data

Visualize data in Light to get an idea of which curve parameters vary for each fish. Look at a random sample of control fish.
```{r, echo=FALSE}
x <- sample(unique(dataL[conc==0,fishID]),12)
ggplot(dataL[fishID%in%x,], aes(x=t,y=y)) +
  geom_point() +
  geom_line() +
  geom_smooth(method = "lm", formula = y ~ poly(x,2), se=FALSE, color="red", lwd = .75) +
  facet_wrap(vars(fishID)) # Intercept seems to be fairly constant
```

It appears that intercept parameter is fairly constant from one fish to another. What appears to vary most are the polynomial parameters for a poly2 fit. Maybe it's fair to say slope varies most.

## Attempt to fit linear mixed effects models to each concentration group individually

### Control parameters for regression fitting.
```{r}
ctrl <- lmeControl(opt = "optim", niterEM = 10000, msTol = 1e-10, msVerbose = TRUE,
                   minAbsParApVar = 1e-10, maxIter = 1000, msMaxIter = 1000,
                   control = list(abstol=1e-10,reltol=1e-10,ndeps=1e-10))
```

### Fit to control data
```{r}
nest0 <- groupedData(y ~ conc|fishID, data =  dataL[conc==0],
                         order.groups = FALSE)
nest0$t <- nest0$t - min(nest0$t)
# Model lme
lmm0 <- lme(y ~ poly(t,degree=2,raw=TRUE),
            correlation = corAR1(, form = ~ t | fishID),
            random = ~ 0 + poly(t,degree=1,raw=TRUE) |fishID,
            method = "ML",
            control = ctrl,
            data = nest0)
lmm0$apVar
```

Model converges.

### Try with next concentration group, 0.004 uM.
```{r}
nest1 <- groupedData(y ~ conc|fishID, data =  dataL[conc==0.004],
                         order.groups = FALSE)
nest1$t <- nest1$t - min(nest1$t)
# Model lme
lmm1 <- lme(y ~ poly(t,degree=2,raw=TRUE),
            correlation = corAR1(, form = ~ t | fishID),
            random = ~ 0 + poly(t,degree=1,raw=TRUE) |fishID,
            method = "ML",
            control = ctrl,
            data = nest1)
lmm1$apVar
```
Fit doesn't work. It is likely that changing some convergence parameters would allow for a non-positive definite matrix, but do we really want to do this? Look at the current estimates for random effects variance of slope for 0.004uM.

```{r, echo = FALSE}
lmm1
```

Standard Deviation is so small, maybe modelling the slope as random for this concentration group is inaccurate. Plot fish trackings for 0.004uM group to investigate this.

```{r, echo = FALSE}
x <- sample(unique(dataL[conc==0,fishID]),12)
ggplot(dataL[conc==0.004], aes(x=t,y=y)) +
  geom_point() +
  geom_line() +
  geom_smooth(method = "lm", formula = y ~ poly(x,2), se=FALSE, color="red", lwd = .75) +
  facet_wrap(vars(fishID)) # Intercept seems to be fairly constant
```

